---
title: "The {sen2rts} workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The {sen2rts} workflow}
  %\VignetteEngine{knitr::knitr}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = TRUE, warning = FALSE,
  fig.width = 9, fig.asp = 0.618, out.width = "100%", fig.align = "center"
)
options(width = 100)
library(sen2r)
library(sen2rts)
samplecrops_dir <- file.path(dirname(attr(load_binpaths(), "path")), "samplecrops")
data(samplecrops)
```

This vignette describes a use case to process a sample Sentinel-2 time series 
and extract information about seasonality.

Functions developed in the package {sen2rts} are run sequentially, following 
this workflow:

1. [produce raster archive with {sen2r}](#produce-raster-archive-with-sen2r);
2. [extract time series over spatial features](#extract-time-series-over-spatial-features);
3. [smooth raw time series and fill gaps](#smooth-raw-time-series-and-fill-gaps);
4. [cut cycles and assign them to specific seasons](#cut-cycles-and-assign-them-to-specific-seasons);
5. [interpolate cycles](#interpolate-cycles);
6. [extract phenological metrics](#extract-phenological-metrics);
7. [aggregate time series over seasons](#aggregate-time-series-over-seasons).


## Produce raster archive with {sen2r}

Package {sen2r} must be used to generate the local raster data over the area of 
interest, since {sen2rts} is built to read a Sentinel-2 image archive in the 
[{sen2r} output structure](https://sen2r.ranghetti.info/articles/outstructure.html). 
In this tutorial the following sample archive is being used:

- **extent**: some field crops of the farm [_Bonifiche Ferraresi_](https://bonificheferraresi.it) 
located in the estate of Jolanda di Savoia (FE, Italy);
- **time window**: 2-years time window (2019 to 2020),
    which can be extended up to 5 years by the user;
- **products**: NDVI 
    [spectral index](https://sen2r.ranghetti.info/articles/outstructure.html#spectral-indices)
    (main product) and 
    [SCL](https://sen2r.ranghetti.info/articles/outstructure.html#scene-classification-map) 
    (for quality metadata).

The following code _could_ be used to produce it 
(but see the note below the code before launching it):

```{r sen2r, eval = FALSE}
library(sen2r)
## Define the directories to store SAFE archives and outputs
samplecrops_dir <- file.path(dirname(attr(load_binpaths(), "path")), "samplecrops")
safe_dir <- file.path(dirname(attr(load_binpaths(), "path")), "safe")
# you can replace the previous lines with the paths of any desired folder
# (in an existing parent directory)
dir.create(samplecrops_dir, showWarnings = FALSE)
dir.create(safe_dir, showWarnings = FALSE)

## Define the output extent
samplecrops_poly <- st_sfc(st_polygon(list(
    matrix(
      c(733500, 4971600,
        733500, 4973580,
        735480, 4973580,
        735480, 4971600,
        733500, 4971600),
      ncol = 2, byrow = TRUE
    )
  )), crs = 32632)

## Install Sen2Cor (used to process SAFE L1C archives)
install_sen2cor()

## Launch sen2r() to download Sentinel-2 data and produce the archive
sen2r(
  gui = FALSE,
  extent = samplecrops_poly,
  extent_name = "samplecrops",
  timewindow = c("2019-01-01", "2020-12-31"),
  list_prods = c("SCL"), # you can add "BOA" for eventual future usage
  list_indices = "NDVI",
  sen2cor_use_dem = TRUE,
  path_l1c = safe_dir,
  path_l2a = safe_dir,
  path_out = samplecrops_dir
)
```
_NOTE:_ producing the output archive with this command requires a lot of patience,
since most of the required SAFE archives must be ordered from 
[ESA Long Term Archive](https://scihub.copernicus.eu/userguide/LongTermArchive)
and Level-1C archives must be processed to obtain the corresponding Level-2A ones.

In order to speed-up this step, the archive can be downloaded from 
this [Zenodo repository](https://doi.org/10.5281/zenodo.4620934)
using the following code instead than the previous chunk:

```{r retrieve_data, eval = FALSE}
## Define the directory to store outputs
samplecrops_dir <- file.path(dirname(attr(load_binpaths(), "path")), "samplecrops")
# you can replace the previous line with the paths of any desired folder
# (in an existing parent directory)
dir.create(samplecrops_dir, showWarnings = FALSE)

## Download and extract the sample archives
samplecrops_url <- "https://zenodo.org/record/4620934/files"
for (sel_y in 2019:2020) { # can be extended up to 2015:2020
  for (sel_p in c("SCL", "NDVI")) {
    download.file(
      file.path(samplecrops_url, paste0("S2X2A_",sel_y,"_022_samplecrops_",sel_p,"_10.zip")),
      temp_zip_path <- tempfile(fileext = ".zip")
    )
    unzip(temp_zip_path, exdir = file.path(samplecrops_dir, sel_p))
    file.remove(temp_zip_path)
  }
}
```

## Extract time series over spatial features

"Raw" time series can be extracted from the data archive using the function
`extract_s2ts()`.
Required inputs are:

1. the paths of the rasters to be used;
2. a spatial file of class `sf` or `sfc` containing the features over which
    extracting the time series.
    
The function `read_s2cube()` can be used to easily obtain the required paths
passing it the main path of the {sen2r} archive, the product type to read 
and some other additional parameters which can be used to filter the existing 
rasters (time window, Sentinel-2 orbits or a specific sensor).
See [the documentation](../reference/read_s2cube.md) for additional details.

The spatial file can contain point or polygon features.
In the first case, raster values overlayed by points are used;
in the second case, cells covered by each polygon are averaged, additionally
using a quality data (SCL and/or CLD) to compute a weighted average.
Moreover, in both the cases the additional quality layers can be used to assign 
a quality flag to each time series value
(see [the function documentation](../reference/extract_sen2rts.md) for details).
In the case of SCL, each surface class can be assigned to a weight value
using the function `scl_weights()`.

In this example, SCL archive is used both to weight averages over polygons
and to assign quality flags to time series.
The default SCL weights are used, with the exception of class "snow"
(since snow is a rare event in the area of interest, pixels classified as snow
are more often errors of classification; so, value "0" is used).

```{r extract_s2ts}
library(sen2rts)
## Read image paths
sen2r_ndvi_paths <- read_s2cube(
  samplecrops_dir, 
  prod_type = "NDVI",
  time_window = c("2019-01-01", "2020-12-31") 
)
sen2r_scl_paths <- read_s2cube(
  samplecrops_dir, 
  prod_type = "SCL",
  time_window = c("2019-01-01", "2020-12-31")
)
# time_window can be extended up to c("2015-07-01", "2020-12-31")

## Read sample polygons
data(samplecrops)
# first 4 polygons are used: skip the following line to use all the 12 ones.
samplecrops <- samplecrops[1:4,]

## Extract TS
ts_ndvi_w <- extract_s2ts(
  sen2r_ndvi_paths, 
  samplecrops, 
  scl_paths = sen2r_scl_paths,
  scl_w = scl_weights(snow = 0),
  in_sf_id = "fid"
)

# Rescale values (this is needed since rasters were saved as Int16 
# with a scale factor of 10000)
ts_ndvi_w$value <- ts_ndvi_w$value / 1E4
```

The output is an object of class `s2ts`, which basically is a `data.table`
associated with specific methods for printing and plotting.
Print method shows a preview of the `data.table`, with first 4 ID and
first/last 5 dates; values are accompanied by symbols which summarises
the value of the quality flag.

```{r print_extract}
print(ts_ndvi_w, topn = 5) # standard visualisation
data.table::as.data.table(ts_ndvi_w) # conversion in data.table allows showing all attributes
plot(ts_ndvi_w)
```

The `plot` method for class `s2ts` calls `ggplot2::ggplot()` (`{ggplot2}`,
which is a suggested dependency, must be installed).
Raw values are shown using a colour scale which depends on the quality flag 
associated to each value.
In this case, it was determined from SCL values, using the following
correspondence between SCL classes and 0-1 values:
```{r scl_weights}
scl_weights(snow = 0)
```

## Smooth raw time series and fill gaps

This step allows: 
1. removing evident temporal discontinuities like drops, usually 
    due to low quality points or points with a wrong associated quality flag
    (misclassification), by applying a Savitzky-Golay low-pass filter;
2. filling gaps (missing dates, or dates characterised by a quality flag = 0)
    to obtain continuous time series.

Function `smooth_s2ts()` is used for the first step.
It takes as input a `s2ts` object created with `extract_s2ts()` (raw time series)
and optional parameter values which can be used to adjust the output
(see the documentation for details), returning a `s2ts` object (the same class 
of the input) containing both raw and smoothed values.
Gap filling can be done with `fill_s2ts()`, with a similar behaviour: it takes
an input smoothed `s2ts` object and setting additional parameters, producing
an output `s2ts` filled time series.

In the following example `smooth_ts()` is launched increasing the default value 
of the low-pass filter, and `fill_s2ts()` with default parameters.

```{r smooth_s2ts}
ts_ndvi_w_s4 <- smooth_s2ts(ts_ndvi_w, sg_daywindow = 15)
ts_ndvi_w_f2 <- fill_s2ts(ts_ndvi_w_s4)
print(ts_ndvi_w_s4, topn = 5)
ts_ndvi_w_f2
plot(ts_ndvi_w_f2)
```
                                                                                                                                                                                                                                                                     
Filled values can be discriminated by smoothed ones because of the
`~` symbol used instead of the quality symbol.

Plotting output time series allows raw values (points
joined by a light-grey line) together with smoothed and filled values
(joined by a dark line).
Points can be hidden with `plot(..., plot_points = FALSE)`
(see the [`plot.s2ts`](../reference/plot.s2ts.md) documentation for details 
about the plotting method).

## Cut cycles and assign them to specific seasons

_Note: here and in the whole package a phenological time window characterised 
by a begin (low index value), a peak (high value) and an end (low value)
is called "cycle".
The term "phenological season" is never used in order to avoid confusion with 
the word "season", used in the package  to identify specific portions of the year
(like in function `assign_season()`).
The only exception is in the definitions of some phenological metrics (e.g. 
Start of Season and End of Season), which are inherited from package {phenopix}
and maintained with the same names._

Next step is the partition of each time series in specific cycles, basing on 
the position of drops (considered as cycle cuts) and peaks.
This is done using the function `cut_cycles()` and properly setting 
arguments `min_win`, `min_peakvalue`, `max_dropvalue` and `min_rel_thresh`:
depending on values of these parameters, each time series is cut in in a higher
or smaller number of cycles.
Moreover, each cycle is associated to a specific year. 
Arguments `n_cycles`, `newyearday` and `weight_metric` allows determining
which year should be associated to each cycle and how many (and which) cycles 
should be maintained for each year.
The selection of cycles to be maintained can be done on the basis of different 
metrics, like the maximum value reached in each cycle (see the 
[function documentation](../reference/cut_cycles.md) for details).
The output is a data table in which each record is a cycle, with associated
dates of begin, end and maximum value; a field `weight` represents the cycle
weight as computed by the function.

In this example, cycles are cut with default parameters with the exception
of the minimum length of the cycle (30 days).

```{r cut_cycles}
cutseas_ndvi_p2 <- cut_cycles(ts_ndvi_w_f2, min_win = 30)
head(cutseas_ndvi_p2)
```

For a finer selection of cycles to be maintained, function `assign_season()`
can be used.
In this case, cycles can be associated not only to a specific year, but also to 
a season (e.g. summer or winter); this is useful for example in case of 
multiple crop seasons.
For each season, the user can define a time windows in which maximum values
must occur, and/or an expected date of the peak to select the nearest cycles.
The function can be also used after having retrieved phenological metrics, 
so to filter on the basis of dates of start or end of season.

In this example the `data.table` produced by `cut_cycles()` is filtered to 
keep a single, summer cycle: in this case, the output data table includes the
same fields of the input.

```{r assign_season}
( cutseas_ndvi_ps1 <- assign_season(
  cutseas_ndvi_p2, max_n_cycles = 1, 
  pop_win = c("04-01", "08-31"), pop_name = "maxval"
) )
```

If the user had asked to keep more than one season (or to assign a name to the 
season), the output would have included an additional field with the season name
(see the example below).

```{r assign_season2}
cutseas_ndvi_ps2 <- assign_season(
  cutseas_ndvi_p2, max_n_cycles = 1, 
  seasons = c("winter", "summer"),
  pop_win = list(c("12-01", "03-31"), c("04-01", "08-31")), 
  pop_name = "maxval"
)
head(cutseas_ndvi_ps2)
```

The retrieved cycles can be plotted together with time series passing
the data table of cycles with the argument `pheno`: in this case,
vertical black lines are drawn to discriminate cycles, and yellow bars
highlight existing cycles (unless `plot_cycles = FALSE`).
Adding the argument `plot_dates = TRUE` allows printing dates of cuts.

```{r plot_assign}
plot(ts_ndvi_w_f2, pheno = cutseas_ndvi_ps1, plot_dates = TRUE, plot_points = TRUE)
```

Notice also that, when `pheno` or `fitted` arguments are passed to the plot,
printing raw points is disabled by default; `plot_points = TRUE` is required
to plot them.

## Interpolate cycles

In order to retrieve phenological metrics, time series have to be interpolated
using a known function.

To perform this step, {sen2rts} makes use of double logistic functions 
implemented in package {phenopix}: `phenopix::PhenoTrs()`, `phenopix::PhenoDeriv()`,
`phenopix::PhenoKlosterman()` and `phenopix::PhenoGu()` (see documentations
for details).
Required inputs are a `s2ts` object generated with `fill_s2ts()` and a data
table of cycles generated with `cut_cycles()`.
Argument `fit` allows choosing the fitting method (multiple methods can be 
passed: in case of failure of the first one, the second is used).
Output is a list containing the interpolations.
since this output is not intended to be visualised by the user but to be passed
to `extract_pheno()`, it can not be easily explored by the user (there are 
not printing methods); 
instead, a plotting method exist, so interpolations can be plotted passing
this object to the argument `fitted` (the example below shows it).
Interpolations are shown with red curves.

```{r fit_curve}
ts_ndvi_cf1 <- fit_curve(ts_ndvi_w_f2, cutseas_ndvi_ps1, fit = "gu")
plot(ts_ndvi_w_f2, fitted = ts_ndvi_cf1)
```

This function can require a large amount of time.
To speed up it, `method = "no"` can be chosen: in this case interpolation is not
performed, and the function simply convert the smoothed and filled time series
in a list in the format accepted by `extract_pheno()`.
User must be aware that extracting phenological metrics from non interpolated series
can lead to errors (generally, only `method = "trs"` works) or wrong estimations.

## Extract phenological metrics

As for fitting methods, also methods to extract phenological metrics are 
inherited from package {phenopix}: `phenopix::PhenoTrs()`, 
`phenopix::PhenoDeriv()`, `phenopix::PhenoKl()` and `phenopix::PhenoGu()`.
Function `extract_pheno()` can be used to do it: it accepts as input a list
of fitted cycles as produced by `fit_curve()` and a method among the available 
ones.
Output is a `data.table` in a format similar to the one produced by
`extract_cycles()`, but containing phenological metrics in addition to dates of
begin and end of cycles.

```{r extract_pheno}
( pheno_ndvi_p5 <- extract_pheno(ts_ndvi_cf1, method = "trs", trs = 0.25) )
plot(ts_ndvi_w_f2, fitted = ts_ndvi_cf1, pheno = pheno_ndvi_p5)
```

Here phenological metrics based on threshold at 10% of relative values is used,
printed and plotted.
Metrics are plotted using segments of different colours.
Different methods produce different metrics; for examples, using 
`method = "klosterman"`:

```{r extract_pheno2}
pheno_ndvi_p6 <- extract_pheno(ts_ndvi_cf1, method = "klosterman")
plot(ts_ndvi_w_f2, pheno = pheno_ndvi_p6)
```

As stated in paragraph 
[Assign season](#cut-cycles-and-assign-them-to-specific-seasons),
season assignment can be also performed after the extraction of metrics.
In the example below, late-seeded crops are extracted basing on dates of 
Start of Season:

```{r assign_after_extract}
( pheno_ndvi_ps1 <- assign_season(
  pheno_ndvi_p5, max_n_cycles = 1, 
  sos_win = c("05-01", "06-15"),
) )
```

##  Aggregate time series over seasons

This additional step can be performed to aggregate time series on the basis 
of the information extracted from the seasonal analysis.

The function accepts as inputs a time series (in the `s2ts` format) and a
data table produced by `cut_cycles()`, `assign_season()` or `extract_pheno()`.
The aggregation can be defined setting arguments `metrics` (the two phenological
metrics to be used as lower and upper limit) and `fun` (the aggregation function).
Output is a `data.table` with the same length of `pheno`, in which each record
contains the aggregate value of the time series over the specified cycle.

In the example below, 95% percentiles among the dates of Start and End of Season
(used as quasi-maximum values) are extracted.

```{r aggregate_pheno}
( ts_ndvi_w_a1 <- aggregate_pheno(
  ts_ndvi_w_f2, pheno_ndvi_ps1, 
  metrics = c("sos", "eos"), fun = "quantile", probs = 0.95, na.rm = TRUE
) )
```

